# Project Name: A BI Exploration of Spotify Songs
# Done by: Joseph Watunu Kamau
# Institutiton: Strathmore University
# Course name: Business Intelligence II


# Dataset
#30000 spotify songs from Joakin Arvidsson on Kaggle.
#data source: https://www.kaggle.com/datasets/joebeachcapital/30000-spotify-songs

# Description
The data this week comes from Spotify via the spotifyr package. Charlie Thompson, Josiah Parry, 
Donal Phipps, and Tom Wolff authored this package to make it easier to get either your own data 
or general metadata arounds songs from Spotify's API. Make sure to check out the 
spotifyr package website to see how you can collect your own data!

# Training vs Test data
We will train and test the model in the ratio 70:30 train:test

# Attributes of the dataset
As provided on Kaggle:
variable class description
track_id character Song unique ID
track_name character Song Name
track_artist character Song Artist
track_popularity double Song Popularity (0-100) where higher is better
track_album_id character Album unique ID
track_album_name character Song album name
track_album_release_date character Date when album released
 playlist_name character Name of playlist -->
 playlist_id character Playlist ID -->
 playlist_genre character Playlist genre -->
 playlist_subgenre character Playlist subgenre -->
 danceability double Danceability describes how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable. -->
 energy double Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy. -->
 key double The estimated overall key of the track. Integers map to pitches using standard Pitch Class notation . E.g. 0 = C, 1 = C♯/D♭, 2 = D, and so on. If no key was detected, the value is -1. -->
 loudness double The overall loudness of a track in decibels (dB). Loudness values are averaged across the entire track and are useful for comparing relative loudness of tracks. Loudness is the quality of a sound that is the primary psychological correlate of physical strength (amplitude). Values typical range between -60 and 0 db. -->
 mode double Mode indicates the modality (major or minor) of a track, the type of scale from which its melodic content is derived. Major is represented by 1 and minor is 0. -->
 speechiness double Speechiness detects the presence of spoken words in a track. The more exclusively speech-like the recording (e.g. talk show, audio book, poetry), the closer to 1.0 the attribute value. Values above 0.66 describe tracks that are probably made entirely of spoken words. Values between 0.33 and 0.66 describe tracks that may contain both music and speech, either in sections or layered, including such cases as rap music. Values below 0.33 most likely represent music and other non-speech-like tracks. -->
 acousticness double A confidence measure from 0.0 to 1.0 of whether the track is acoustic. 1.0 represents high confidence the track is acoustic. 
 instrumentalness double Predicts whether a track contains no vocals. "Ooh" and "aah" sounds are treated as instrumental in this context. Rap or spoken word tracks are clearly "vocal". The closer the instrumentalness value is to 1.0, the greater likelihood the track contains no vocal content. Values above 0.5 are intended to represent instrumental tracks, but confidence is higher as the value approaches 1.0.
 liveness double Detects the presence of an audience in the recording. Higher liveness values represent an increased probability that the track was performed live. A value above 0.8 provides strong likelihood that the track is live.
 valence double A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry).
 tempo double The overall estimated tempo of a track in beats per minute (BPM). In musical terminology, tempo is the speed or pace of a given piece and derives directly from the average beat duration.
 duration_ms double Duration of song in milliseconds 

First we load the dataset

```{r}

library(readr)
library(dplyr)

spotify_sampled_data<-read_csv(
  "data/spotify_songs.csv",
  col_types = cols(
    track_id = col_character(),
    track_name = col_character(),
    track_artist = col_character(),
    track_popularity = col_double(),
    track_album_id = col_character(),
    track_album_name = col_character(),
    track_album_release_date = col_character(),
    playlist_name = col_character(),
    playlist_id = col_character(),
    playlist_genre = col_factor(levels = c("edm", "pop","latin","r&b", "rap","rock")),
    playlist_subgenre = col_character(),
    danceability = col_double(),
    energy = col_double(),
    key = col_double(),
    loudness = col_double(),
    mode = col_factor(levels= c("0", "1")),
    speechiness = col_double(),
    acousticness = col_double(),
    instrumentalness = col_double(),
    liveness = col_double(),
    valence = col_double(),
    tempo = col_double(),
    duration_ms = col_double()
  )
)

```

Due to current processing capability, we will reduce the dataset to 10000 columns
```{r}
spotify_sampled_data <- spotify_sampled_data %>% 
  sample_n(10000)


View(spotify_sampled_data)

dim(spotify_sampled_data)
```

We will now Identify the data types in the datatset. ThiS will help us choose the
most efficient visualization types and algorithms that we can use.
```{r}
sapply(spotify_sampled_data, class)
```

Now we will perform descriptive statistics.


# 1.Measure of frequency.
 Here we will try to understand number of instances that belong to each class.
```{r}
song_data_genre_freq<- spotify_sampled_data$playlist_genre
cbind(frequency=table(song_data_genre_freq),
      percentage=prop.table(table(song_data_genre_freq))*100)

song_data_mode_freq <- spotify_sampled_data$mode
cbind(frequency=table(song_data_mode_freq),
                 percentage=prop.table(table(song_data_mode_freq))*100)
```


```{r}

song_data_subgenre_freq <- spotify_sampled_data$playlist_subgenre
cbind(frequency=table(song_data_subgenre_freq),
      percentage=prop.table(table(song_data_subgenre_freq))*100)

```

# Calculating the mode

```{r}
song_data_genre_mode <- names(table(spotify_sampled_data$playlist_genre))[
  which(table(spotify_sampled_data$playlist_genre)==max(table(spotify_sampled_data$playlist_genre)))
]
print(song_data_genre_mode)


song_data_subgenre_mode <- names(table(spotify_sampled_data$playlist_subgenre))[
  which(table(spotify_sampled_data$playlist_subgenre)==max(table(spotify_sampled_data$playlist_subgenre)))
]
print(song_data_subgenre_mode)


song_playlistname_mode <- names(table(spotify_sampled_data$playlist_name))[
  which(table(spotify_sampled_data$playlist_name)==max(table(spotify_sampled_data$playlist_name)))
]
print(song_playlistname_mode)
```

# Measuring the distribution of the data for each variable
```{r}
summary(spotify_sampled_data)
```

# Measuring the standard deviation of each variable
Important point to note is Low variability is ideal because it means that you
can predict information about the population based on sample data. High variability
means that the values are less consistent, thus making it harder to make predictions.

```{r}
sapply(spotify_sampled_data[, c(4,12,13,14,15,17,18,19,20,21,22,23)], class)

sapply(spotify_sampled_data[, c(4,12,13,14,15,17,18,19,20,21,22,23)], sd)
```

# Meassuring the variance of each variable

```{r}
sapply(spotify_sampled_data[, c(4,12,13,14,15,17,18,19,20,21,22,23)], var)
```

# Measuring the Kurtosis of each variable
Kurtosis informs you of how often outliers occur in the results

First we will install the necessary packages

```{r}

if (!is.element("e1071", installed.packages()[, 1])) {
  install.packages("e1071", dependencies = TRUE)
}
require("e1071")
```

```{r}
sapply(spotify_sampled_data[, c(4,12,13,14,15,17,18,19,20,21,22,23)], kurtosis, type=2)
```

# Measuring the skewness of each variable

```{r}
sapply(spotify_sampled_data[, c(4,12,13,14,15,17,18,19,20,21,22,23)], skewness, type=2)
```
# Measuring the relationships

```{r}
songs_data_cov <- cov(spotify_sampled_data[, c(4,12,13,14,15,17,18,19,20,21,22,23)])
View(songs_data_cov)
```

# Measuring correlation between variables
```{r}

song_data_cor <- cor(spotify_sampled_data[,c(4,12,13,14,15,17,18,19,20,21,22,23)])
View(song_data_cor)
```

# One way ANOVA to see if there are significant differences in the means of track popularity across different genres

```{r}
song_data_oneway_anova <- aov(track_popularity ~ playlist_genre, data = spotify_sampled_data)
summary(song_data_oneway_anova)
```

# Two-way ANOVA to test the impact of both playlist genre and playlist subgenre on  danceability of a song

```{r}
song_data_twoway_anova<- aov(danceability ~ playlist_genre* playlist_subgenre, data= spotify_sampled_data)
summary(song_data_twoway_anova)
```

# Two-way ANOVA to test the impact of both playlist genre and playlist subgenre on  popularity of a song

```{r}
song_data_twoway_anova<- aov(track_popularity ~ playlist_genre* playlist_subgenre, data= spotify_sampled_data)
summary(song_data_twoway_anova)
```

# Two-way ANOVA to test the impact of both playlist genre and playlist subgenre on  acousticness of a song

```{r}
song_data_twoway_anova<- aov(acousticness ~ playlist_genre* playlist_subgenre, data= spotify_sampled_data)
summary(song_data_twoway_anova)
```


# Basic visualization to understand the dataset
Histogram to represent the dataset

```{r}

par(mfrow = c(1, 3))  

boxplot(spotify_sampled_data[, 21], main = names(spotify_sampled_data)[21])
boxplot(spotify_sampled_data[, 23], main = names(spotify_sampled_data)[23])
boxplot(spotify_sampled_data[, 22], main = names(spotify_sampled_data)[22])
boxplot(spotify_sampled_data[, 20], main = names(spotify_sampled_data)[20])
boxplot(spotify_sampled_data[, 19], main = names(spotify_sampled_data)[19])
boxplot(spotify_sampled_data[, 18], main = names(spotify_sampled_data)[18])
boxplot(spotify_sampled_data[, 17], main = names(spotify_sampled_data)[17])
boxplot(spotify_sampled_data[, 16], main = names(spotify_sampled_data)[16])
```

Creating a map to identify the missing data in each dataset
loading necessary datasets
```{r}
if (!is.element("Amelia", installed.packages()[, 1])) {
  install.packages("Amelia", dependencies = TRUE)
}
require("Amelia")
missmap(spotify_sampled_data, col = c("red", "grey"), legend = TRUE)


if (!is.element("corrplot", installed.packages()[, 1])) {
  install.packages("corrplot", dependencies = TRUE)
}
require("corrplot")
```

We now create a correlation plot

```{r}
corrplot(cor(spotify_sampled_data[c(4,12,13,14,15,17,18,19,20,21,22,23)]), method = "circle")
```

We will now perform data Imputation. We will try and check for missing values and
 try to remove them for data consistency.

First we will install the necessary packages to perform imputation

```{r}

if(!is.element("NHANES", installed.packages()[, 1])){
  install.packages("NHANES", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

require("NHANES")


if(!is.element("dplyr", installed.packages()[, 1])){
  install.packages("dplyr", dependencies=TRUE,
                   repos = "https://cloud.r-project.org")
}

require("dplyr")


if(!is.element("naniar", installed.packages()[, 1])){
  install.packages("naniar", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

require("naniar")


if(!is.element("mice", installed.packages()[, 1])){
  install.packages("mince", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

require("mice")


if(!is.element("Amelia", installed.packages()[, 1])){
  install.packages("Amelia", dependencies = TRUE,
                   repos="https://cloud.r-project.org")
}

require("Amelia")

```

We now confirm if there are any missing value
```{r}
any_na(spotify_sampled_data)
```
We check how many are missing

```{r}
n_miss(spotify_sampled_data)
```

For clarity, we check what percentage of missing data is in the entire dataset

```{r}
prop_miss(spotify_sampled_data)
```

How many missing values does each variable have

```{r}
spotify_sampled_data %>% is.na() %>% colSums()
```

Checking the number and percentage of missing values grouped by each observation

```{r}
miss_var_summary(spotify_sampled_data)

miss_case_summary(spotify_sampled_data)
```

Checking which variable contain the most missing values

```{r}
gg_miss_var(spotify_sampled_data)
```

which combinations fo variables are missing together

```{r}
gg_miss_upset(spotify_sampled_data)
```

creating a heatmap of missingness

```{r}
is.factor(spotify_sampled_data$track_name)
```


Now we will perfrom data Transformation necessary for improving the accuracy of 
the final model.

```{r}
if(!is.element("caret", installed.packages()[, 1])){
  install.packages("caret", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")

}
require("caret")
```


```{r}

model_of_the_transform <- preProcess(spotify_sampled_data, method=c("scale"))
print(model_of_the_transform)

spotify_songs_scale_transform <- predict(model_of_the_transform, spotify_sampled_data)

summary(spotify_songs_scale_transform)
```

# Center Data Transform

Before

```{r}
summary(spotify_sampled_data)

boxplot(spotify_sampled_data[, 21], main = names(spotify_sampled_data)[21])
boxplot(spotify_sampled_data[, 23], main = names(spotify_sampled_data)[23])
boxplot(spotify_sampled_data[, 22], main = names(spotify_sampled_data)[22])
boxplot(spotify_sampled_data[, 20], main = names(spotify_sampled_data)[20])
boxplot(spotify_sampled_data[, 19], main = names(spotify_sampled_data)[19])
boxplot(spotify_sampled_data[, 18], main = names(spotify_sampled_data)[18])
boxplot(spotify_sampled_data[, 17], main = names(spotify_sampled_data)[17])
boxplot(spotify_sampled_data[, 16], main = names(spotify_sampled_data)[16])
boxplot(spotify_sampled_data[, 15], main = names(spotify_sampled_data)[15])
boxplot(spotify_sampled_data[, 14], main = names(spotify_sampled_data)[14])
boxplot(spotify_sampled_data[, 13], main = names(spotify_sampled_data)[13])
boxplot(spotify_sampled_data[, 12], main = names(spotify_sampled_data)[12])

```
```{r}
model_of_the_transform <- preProcess(spotify_sampled_data, method = c("center"))
print(model_of_the_transform)
spotify_data_center_transform<- predict(model_of_the_transform, # nolint
                                           spotify_sampled_data)
```


After
```{r}

summary(spotify_data_center_transform)
boxplot(spotify_data_center_transform[, 21], 
        main = names(spotify_data_center_transform)[21])
boxplot(spotify_data_center_transform[, 23], 
        main = names(spotify_data_center_transform)[23])
boxplot(spotify_data_center_transform[, 22], 
        main = names(spotify_data_center_transform)[22])
boxplot(spotify_data_center_transform[, 20], 
        main = names(spotify_data_center_transform)[20])
boxplot(spotify_data_center_transform[, 19], 
        main = names(spotify_data_center_transform)[19])
boxplot(spotify_data_center_transform[, 18], 
        main = names(spotify_data_center_transform)[18])
boxplot(spotify_data_center_transform[, 17], 
        main = names(spotify_data_center_transform)[17])
boxplot(spotify_data_center_transform[, 16], 
        main = names(spotify_data_center_transform)[16])
boxplot(spotify_data_center_transform[, 15], 
        main = names(spotify_data_center_transform)[15])
boxplot(spotify_data_center_transform[, 14], 
        main = names(spotify_data_center_transform)[14])
boxplot(spotify_data_center_transform[, 13], 
        main = names(spotify_data_center_transform)[13])
boxplot(spotify_data_center_transform[, 12], 
        main = names(spotify_data_center_transform)[12])
```

# Standadize data transformation
Before
```{r}
summary(spotify_sampled_data)
sapply(spotify_sampled_data[, c(4,12,13,14,15,17,18,19,20,21,22,23)], sd)

model_of_the_transform<- preProcess(spotify_sampled_data,
                                    method = c("scale", "center"))

print(model_of_the_transform)
spotify_data_standardize_transform<-predict(model_of_the_transform,
                                           spotify_sampled_data)
```

After

```{r}
summary(spotify_data_standardize_transform)
sapply(spotify_data_standardize_transform[, c(4,12,13,14,15,17,18,19,20,21,22,23)], sd)
```

# Normalize data Transform

```{r}
summary(spotify_sampled_data)
model_of_the_transform <- preProcess(spotify_sampled_data, method = c("range"))
print(model_of_the_transform)
spotify_data_normalize_transform <- predict(model_of_the_transform,
                                            spotify_sampled_data)

summary(spotify_data_normalize_transform)
```

# Splitting the dataset in a 75:25 train:test ratio

```{r}
str(spotify_sampled_data)

train_index <- createDataPartition(spotify_sampled_data$playlist_genre,p=0.75,list=FALSE)

spotify_data_train<- spotify_sampled_data[train_index, ]
spotify_data_test<- spotify_sampled_data[-train_index, ]
```

